Normalization by Evaluation for Call-by-Push-Value and Polarized Lambda-Calculus

### 2019-02-10 fscd19/fscd19.tex
[ fscd19 ] Terms of CBPV

Defining%
\footnote{%
A more accurate presentation of $\tNF$ would be
$\sum_{u : \tNf A\,\Gamma} (\tm u \cong t)$
in order to make absolutely clear that we
want get hold of the witness $u$ of the existential statement;
however, we shall only use constructive
existentials
in this article and thus stick to the notation $\exists$.
We mean \emph{constructive existence} in
  contrast to classical existence and
  \emph{mere existence} from Homotopy Type Theory which
  requires propositional truncation.
}
\fbox{$\NF t = \exists n \of \tNf A\,\Gamma.\ \tm n \cong t$}
% for $t : A \from \Gamma$,
we aim to construct a normalization function
\[
  \tnorm : (t : A \from \Gamma) \to \NF t
\]
that gives us for each term $t$ a $\cong$-equal normal form $n$.

@@

>>>>

Defining%
\footnote{%
A more accurate presentation of $\tNF$ would be
$\Sigma\,{u : \tNf A\,\Gamma}.\ (\tm u \cong t)$
in order to make absolutely clear that we
want get hold of the witness $u$ of the existential statement;
however, we shall only use constructive
existentials
in this article and thus stick to the notation $\exists$.
We mean \emph{constructive existence} in
  contrast to classical existence and
  \emph{mere existence} from Homotopy Type Theory which
  is the propositional truncation of constructive existence.
}
\fbox{$\NF t = \exists n \of \tNf A\,\Gamma.\ \tm n \cong t$}
% for $t : A \from \Gamma$,
we aim to construct a normalization function
\[
  \tnorm : (t : A \from \Gamma) \to \NF t
\]
that gives us for each term $t$ a $\cong$-equal normal form $n$.

%Equational theory \fbox{$t \cong t'$} for $t,t' : A \from \Gamma$.
Term equality\fbox{$t \cong t'$} for $t,t' : A \from \Gamma$ is the
least congruence over the following axioms, which we have grouped into
$\beta$ (computation), $\eta$ (extensionality), and $\pi$ (permutation)
rules.  We may refer to subgroups, like $\eta^N$ for the negative
$\eta$-rules ($\retato$, $\retatimes$, $\retaone$).  Note that all
typing restrictions are implicit in the well-typedness assumption,
\eg, $t \cong \tunit$ presupposes $t : \tyone \from \Gamma$
since $\tunit : \tyone \from \Gamma$.
\begin{gather*}
%% BETA LAWS
  \nru{\rbetato}{}{\tapp\,(\tabs\,t)\,u \cong \subone t u}
\qquad
  \nru{\rbetatimes}{}{\prjP i {\tpair\,t_1\,t_2} \cong t_i}
\qquad
  \nru{\rbetaplus}{}{\tcase\,(\inj i t)\,t_1\,t_2 \cong t_i}
\\[2ex]
%% ETA LAWS
  \nru{\retato}{}{t \cong
    \tabs\,(\tapp\,(\tren\,\twk\,t)\,(\tvar\,\tzero))}
\quad
  \nru{\retatimes}{}{t \cong \tpair\,(\prj 1 t)\,(\prj 2 t)}
\quad
  \nru{\retaone}{}{t \cong \tunit}
\\[1ex]
  \nru{\retaplus}{}{t \cong
    \tcase\,t
      \,(\injP 1 {\tvar\,\tzero})
      \,(\injP 2 {\tvar\,\tzero})
    }
\qquad
  \nru{\retazero}{}{t \cong \tabort\,t}
\\[2ex]
%% PERMUTATIONS
  \nru{\rpi{\To}{\tyempty}}
    {}{\tapp\,(\tabort\,t)\,u \cong \tabort\,t}
\qquad
  \nru{\rpi{\To}{+}}
      {t_i' = \tapp\,t_i\,(\tren\,\twk\,u)}
      {\tapp\,(\tcase\;t\;t_1\;t_2)\,u \cong \tcase\;t\;t_1'\;t_2'}
\\[1ex]
  \nru{\rpi{\times}{\tyempty}}
    {}{\prjP i {\tabort\,t} \cong \tabort\,t}
\qquad
  \nru{\rpi{\times}{+}}
    {}{\prjP i {\tcase\;t\;t_1\;t_2} \cong
         \tcase\,t\,(\prj i {t_1})\,(\prj i {t_2})}
\\[2ex]
  \nru{\rpi{+}{\tyempty}}
    {}{\tcase\,(\tabort\,t)\,t_1\,t_2 \cong \tabort\,t}
\qquad
  \nru{\rpi{+}{+}}
      {t_i' = \tcase\;t_i\;u_1'\;u_2'
       \quad
       u_j' = \tren\,(\tlift\,\twk)\,u_j
      }
      {\tcase\,(\tcase\;t\;t_1\;t_2)\,u_1\,u_2 \cong
        \tcase\;t\;t_1'\;t_2'}
    \
    % t_i' = \tcase\;t_i
    %   \,(\tren\,(\tlift\,\twk)\,u_1)
    %   \,(\tren\,(\tlift\,\twk)\,u_2)
\\[1ex]
  \nru{\rpi{\tyempty}{\tyempty}}
    {}{\tabort\,(\tabort\,t) \cong \tabort\,t}
\qquad
  \nru{\rpi{\tyempty}{+}}
    {}{\tabort\,(\tcase\;t\;t_1\;t_2) \cong
      \tcase\,t\,(\tabort\,t_1)\,(\tabort\,t_2)}
%% OLD:
%   \nru{\rpi{\To}{\tyempty}}
%     {}{\tapp\,(\tabort\,t)\,u \cong \tabort\,t}
% \qquad
%   \nru{\rpi{\times}{\tyempty}}
%     {}{\prjP i {\tabort\,t} \cong \tabort\,t}
% \\[1ex]
%   \nru{\rpi{+}{\tyempty}}
%     {}{\tcase\,(\tabort\,t)\,t_1\,t_2 \cong \tabort\,t}
% \qquad
%   \nru{\rpi{\tyempty}{\tyempty}}
%     {}{\tabort\,(\tabort\,t) \cong \tabort\,t}
% \\[2ex]
%   \nru{\rpi{\To}{+}}
%     {}{\tapp\,(\tcase\;t\;t_1\;t_2)\,u \cong \tcase\;t\;t_1'\;t_2'}
%     \
%     t_i' = \tapp\,t_i\,(\tren\,\twk\,u)
%   % \nru{\rpi{\To}{+}}
%   %   {}{\tapp\,(\tcase\;t\;t_1\;t_2)\,u \cong \tcase\,t
%   %        \, (\tapp\,t_1\,(\tren\,\twk\,u))
%   %        \, (\tapp\,t_2\,(\tren\,\twk\,u))
%   %   }
% \\
%   \nru{\rpi{\times}{+}}
%     {}{\prjP i {\tcase\;t\;t_1\;t_2} \cong
%          \tcase\,t\,(\prj i {t_1})\,(\prj i {t_2})}
% \\
%   \nru{\rpi{+}{+}}
%     {}{\tcase\,(\tcase\;t\;t_1\;t_2)\,u_1\,u_2 \cong
%         \tcase\;t\;t_1'\;t_2'}
%     \
%     t_i' = \tcase\;t_i\;u_1'\;u_2'
%     ; \quad
%     u_j' = \tren\,(\tlift\,\twk)\,u_j
%     % t_i' = \tcase\;t_i
%     %   \,(\tren\,(\tlift\,\twk)\,u_1)
%     %   \,(\tren\,(\tlift\,\twk)\,u_2)
% \\
%   \nru{\rpi{\tyempty}{+}}
%     {}{\tabort\,(\tcase\;t\;t_1\;t_2) \cong \tcase\,t\,(\tabort\,t_1)\,(\tabort\,t_2)}
\end{gather*}
The different law classes contribute to the shape and completeness of
the normal forms in the following way:
Thanks to the $\eta$-laws, a normal form of negative type is always an
introduction.  A normal form of sum type ($+$, $\tyzero$) is always a
case tree whose leaves are injections.  A normal form of atomic type
($o$) is always case tree whose leaves are neutral.
The $\beta$ and $\pi$ laws determine the shape of neutrals.
Thanks to the $\beta$-laws,
the principal argument of normal eliminations cannot be an
introduction.  Because of the negative permutation rules
$\rpi N P$ it cannot be positive elimination
($\tcase$, $\tabort$) either, which leaves only negative eliminations
in neutrals ($\tapp$, $\tprj$).  Finally the positive permutation
rules $\rpi P P$ guarantee that having neutral scrutinees in case
trees is sufficient.

@@

\[
\begin{array}{lllrl@{\qquad}l}
  \PTy & \ni & P,Q & ::= & \patom
    \mid \tyzero \mid P_1 + P_2
    \mid \tyone \mid P_1 \times P_2
    \mid \Thunk N
  & \mbox{positive type}
\\
  \NTy & \ni & N,M & ::= & \natom
    \mid \top \mid N_1 \ntimes N_2
    \mid P \To N
    \mid \Comp P
  & \mbox{negative type}
\\
  \HTy & \ni & H & ::= & \patom \mid P
  & \mbox{hypothesis type}
\\
\end{array}
\]

Pure CBPV adds right non-invertible focusing.

Assuming a family $\T\,N\,\Gamma$ of terms of negative type $N$ in
context $\Gamma$, values \fbox{$v : \PValT P \Gamma$}
of positive type $P$ are constructed by the
following rules:
\begin{gather*}
  \nru{\pvar}{P \in \Gamma}{\PValT P \Gamma}
\qquad
  \nru{\tthunk}{\T\,N\,\Gamma}{\PValTP{\Thunk N}\Gamma}
\\[2ex]
  \nru{\punit}{}{\PValT{\tyone}\Gamma}
\qquad
  \nru{\ppair}
      {\PValT{P_1}\Gamma \qquad \PValT{P_2}\Gamma}
      {\PValTP{P_1 \times P_2}\Gamma}
\qquad
  \nru{\tinj_i}{\PValT{P_i}\Gamma}{\PValTP{P_1 + P_2}\Gamma}
\end{gather*}

It relies on the previous definitions
with instance $\T = \tNf$.
\fbox{$\Tm N \Gamma$}:
\begin{gather*}
  \nru{\tret}
      {\PVal P \Gamma}
      {\TmP {\Comp P} \Gamma}
\quad
  \nru{\tabs}
      {\Tm N {(\cext \Gamma P)}}
      {\TmP {P \To N} \Gamma}
\quad
  \nru{\npair}
      {\Tm {N_1} \Gamma \qquad \Tm {N_2} \Gamma}
      {\TmP {N_1 \ntimes N_2} \Gamma}
\quad
  \nru{\nunit}
      {}
      {\Tm \top \Gamma}
\\[2ex]
  \nru{\tforce}
      {\PValP {\Thunk N} \Gamma}
      {\Tm N \Gamma}
\qquad
  \nru{\tapp}
      {\TmP {P \To N} \Gamma \qquad \PVal P \Gamma}
      {\Tm N \Gamma}
\qquad
  \nru{\tprj_i}
      {\TmP {N_1 \ntimes N_2} \Gamma}
      {\Tm {N_i} \Gamma}
\\[2ex]
% \end{gather*}
% \begin{gather*}
  \nru{\tbind}
      {\TmP {\Comp P} \Gamma \qquad \Tm N {(\cext \Gamma P)}}
      {\Tm N \Gamma}
\qquad
  \nru{\tsplit}
    {\TmP {P_1 \times P_2} \Gamma
      \qquad \Tm N {(\cext {\cext \Gamma P_1} P_2)}
    }{\Tm N \Gamma}
\\[2ex]
  \nru{\tcase}
    {\TmP {P_1 + P_2} \Gamma
      \qquad \Tm N {(\cext \Gamma P_1)}
      \qquad \Tm N {(\cext \Gamma P_2)}
    }{\Tm N \Gamma}
\qquad
  \nru{\tabort}
    {\Tm \tyzero \Gamma}
    {\Tm N \Gamma}
\end{gather*}
